---
title: descstat, an R Package for Computing Descriptive Statistics
author: Yves Croissant
date: 2020/11/29
output: 
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    number_sections: true

vignette: >
  %\VignetteIndexEntry{desctat: an R package for descriptive statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      out.width = '70%', fig.asp = 0.5,
                      fig.align = "center")
options(
    htmltools.dir.version = FALSE, formatR.indent = 2, width = 55, digits = 4,
    tibble.print_max = 5, tibble.print_min = 5)
otheme <- ggplot2::theme_set(ggplot2::theme_minimal())
```

**R** offers many tools to analyse the univariate or bivariate
distribution of series. This includes `table` and `prop.table` on base
**R**, and `group_by/summarise` and `count` in **dplyr**. However,
these functions are somehow frustrating as some very common tasks,
like:

- adding a total,
- computing relative frequencies or percentage instead of counts,
- merging high values of integer series in a `>=N` category,
- computing cumulative distributions, 

are tedious. Moreover, to our knowledge, **R** offers weak support
for numerical series for which the numerical value is not known at the
individual level, but only the fact that this value belongs to a
certain class. `descstat` is intended to provide user-friendly tools
to perform these kind of operations. More specifically, three kind of
tables can be constructed with `descstat`:

- `freq_table` for frequency tables, suitable for factors or integer
  numerical series,
- `bins_table` for bins tables, suitable for numerical series,
  either provided as a numeric or as a factor containing numerical
  classes,
- `cont_table` for contingency tables of two series.

These function are writen in the **tidyverse** style, which means that
the pipe operator can be used and that the series can be selected
without quotes.

```{r }
library("ggplot2")
library("descstat")
```

# Frequency tables

Frequency tables are suitable to summarize the univariate distribution
of a categorical or an integer series. In `tidyverse`, this task can be
performed using the `dplyr::count` function. For example, the `rgp`
data set, which is an extract of the French cencus, contains the
number of children in households:

```{r }
rgp %>% count(children)
```

## Computing a frequency table

The `descstat::freq_table` function performs the same task:

```{r }
rgp %>% freq_table(children)
```

but it has several further arguments which can improve the result:

- `cols` is a character containing one or several of the following
  letters:
  
    - `n` for the count or absolute frequencies (the default),
    - `f` for the (relative) frequencies,
    - `p` for the percentage (*ie* $f\times 100$),
	- `N`, `F` and `P` returns the cumulative values of `n`, `f` and
      `p`.
- `total` a boolean, if `TRUE`, a total is returned,
- `max`, suitable for an integer series only, is an integer which is the
  maximum value presented in the frequency table, *eg* `max = 3` creates
  a last line which is `>=3`.
  
The following command use all the possible letters.

```{r }
rgp %>% freq_table(children, cols = "nfpNFP")
```

As there are few occurrences of families with more than 3 children, we
set `max = 3` and add a total by setting `total` to `TRUE`.

```{r }
rgp %>% freq_table(children, max = 3, total = TRUE)
```

## Printing a frequency table

Note that in the printed table, the `children` series is now a
character as the last two values are `>=3` and `Total`. Actually
`freq_table` returns an object of class `freq_table` which inherits
from the `tbl_df` class. A look at the structure of the object:

```{r }
rgp %>% freq_table(children, max = 3, total = TRUE) %>% str
```

indicates that we still have a tibble, with a numeric `children`
series for which the last two values equal to 3.5 (for 3 and more) and
`NA` (for the total).

A `pre_print` function is provided with a method for `freq_table`
objects. It turns the `children` series in a `character`, with `3.5`
and `NA` replaced by `>=3` and `Total`. This `pre_print` method is
included in the `format` method for `freq_table` objects, which is then
passed to the `tbl_df` method:

```{r }
descstat:::format.freq_table
```
The `pre_print` function should be used explicitly while using
`knitr::kable`, as this function doesn't use any `format` method:

```{r }
rgp %>% freq_table(children, max = 3, total = TRUE) %>%
    pre_print %>% knitr::kable()
```

## Ploting a frequency table

The most natural way to plot a frequency table with `ggplot` is to use
`geom_col` (or equivalently `geom_bar` with `stat = 'identity'`).

```{r }
cld <- rgp %>% freq_table(children, cols = "nf", max = 3)
cld %>% pre_print %>% ggplot(aes(children, f)) +
    geom_col(fill = "white", color = "black")
```

Note the use of the `pre_print` method which turns the `3.5` numerical
value in `>=3`.

To get more enhanced graphics, a `pre_plot` method is provided, with a
`plot` argument equal to `banner` or `cumulative`. With `plot =
"banner"`,

```{r }
cld %>% pre_print %>% pre_plot("f", plot = "banner")
```

`pre_plot` returns an `ypos` series which indicates the coordinates
where to write the label corresponding to the frequencies.

```{r }
bnp <- cld %>% pre_print %>% pre_plot("f", plot = "banner") %>%
    ggplot(aes(x = 2, y = f, fill = children)) +
    geom_col() +
    geom_text(aes(y = ypos, label = f)) +
    scale_x_continuous(label = NULL) +
    scale_fill_brewer(palette = "Set3")
bnp
```

using polar coordinates, we get a pie chart:

```{r }
bnp + coord_polar(theta = "y") + theme_void()
```

changing the range of the `x` values, we get a hole in the pie chart
which result in the so called donut chart:

```{r }
bnp + scale_x_continuous(limits = c(1, 2.5)) +
    coord_polar(theta = "y") + theme_void()
```

The other possible value for the `plot` argument of `pre_plot` is
cumulative:


```{r }
cld <- rgp %>% freq_table(children, "F", max = 5, total = TRUE)
cld %>% pre_plot(plot = "cumulative") %>% print(n = 5)
```
this returns four series which have the names of the aesthetics that
are mandatory for `geom_segment`; `x`, `xend`, `y` and `yend`. It
also returns a `pos` series with which one can draw differently the
horizontal (`hor`) and the vertical (`vert`) segments, using for
example the `linetype` aesthetic.


```{r }
cld %>% pre_plot(plot = "cumulative") %>% 
    ggplot() +
    geom_segment(aes(x = x, xend = xend, y = y, yend = yend,
                     linetype = pos)) +
    guides(linetype = FALSE) +
    labs(x = "number of children", y = "cumulative frequency")
```

# Bins tables


`descstat::bins_table` computes a bins table, *ie* a table that
contains the frequencies for different classes of a numerical
series. This numerical series can either be numerical or coded as a
class in the original (raw) tibble.

For example, the `wages` data set contains two series called `wage`
and `size` which respectively indicate the class of wages and the
class of firm size.

```{r }
wages %>% print(n = 3)
```

The univariate distribution of, for example, `size` can be computed using the
`dplyr::count` or the `descstat::freq_table` functions.


```{r }
wages %>% count(size)
wages %>% freq_table(size)
```

## Creating bins tables

The `bins_table` function provides a richer interface. Firstly, it
returns a series called `x` which is the center of the
classes. Secondly, a `break` argument is provided, which is a
numerical vector that can be used to reduce the number of classes:

```{r }
wages %>% bins_table(size, breaks = c(20, 250))
```

or to create classes from numerical values:

```{r }
padova %>% pull(price) %>% range
padova %>% bins_table(price, breaks = c(250, 500, 750))
padova %>% bins_table(price, breaks = c(30, 250, 500, 750, 1000))
```

Note that in this latter case, the first (last) values of `breaks`
can be either:

- outside the range of the series; in this case the lower bound of the
  first class and the upper bound of the last class are these values,
- inside the range of the series; in this case the lower bound of the
  first class is `0` and the upper bound of the last class is `Inf`. 

Moreover, as for `freq_table`, the `cols` argument enables the
computation of counts, frequencies and percentage and the cumulative
values are obtained using upper caps.

```{r }
wages %>% bins_table(size, cols = "nFp", breaks = c(20, 100, 250))
```

`cols` can also contain a `d` for density, which is the frequency
divided by the class width, `m` which is the mass for the class (the
product of the frequency and the value of the variable) and `M` which
is the cumulated mass.

```{r }
wages %>% bins_table(size, cols = "dmM", breaks = c(20, 100, 250))
```

A total can be computed for `f`, `n` and `p`, as for `freq_table`
objects.

```{r }
wages %>% bins_table(size, cols = "dMfnFp", breaks = c(20, 100, 250),
                     total = TRUE)
```

## Classes and values

For the computation of descriptive statistics, classes should be
replaced by values. By the default, the center of the class is used,
which means that the computation is done as if all the observations of
the class had a value equal to its center. This value is returned as
`x` while using `bins_table`. Three arguments control what happens for
the first and the last class:

- a specific value for the first class can be indicated using the
  `xfirst` argument,
- a specific value for the last class can be indicated using the
  `xlast` argument,
- the width of the last class (if it is opened on infinity on the right)
  can be set as a multiple of the before last class using the
  `wlast` argument. 

To set the center of the first class to `10` (and not to `10.5`), we
can use:

```{r }
wages %>% bins_table(size, breaks = c(20, 100, 250), xfirst = 10)
```

To set the center of the last class to `400` we can either set
`xlast` to this value or to indicate that the width of the last class
should be twice the one of the before last:

```{r results = 'hide'}
wages %>% bins_table(size, breaks = c(20, 100, 250), xlast = 400)
```

```{r }
wages %>% bins_table(size, breaks = c(20, 100, 250), wlast = 2)
```

other values of the variable can be included in the table using the
`vals` argument, which is a character including some of the following
letters:

- `l` for the lower bound of the class,
- `u` for the upper bound of the class,
- `a` for the width of the class.

```{r }
wages %>% bins_table(size, vals = "xlua", cols = "p", breaks = c(20, 100, 250), wlast = 2)
```

## Under the hood: converting classes to values

`bins_table` internally calls the function `cls2val` which turns the
classes into values of the variable. Its main argument is a factor (or
a character) containing classes of numerical values, *ie* a string of
the form `[a,b)` or `(a,b]` where `a` and `b` are respectively the
lower and the upper values of the classes. In the first case, the
class is closed on the left (`a` is included) and opened on the right
(`b` is not included) as for the second case, the class is opened on
the left and closed on the right. This is the notation used when the
`base::cut` function is used to transform a numerical value to a
class, using a vector of breaks.

```{r }
padova %>% pull(price) %>% head
padova %>% pull(price) %>%
    cut(breaks = c(0, 250, 500, 1000), right = FALSE) %>% head
```

Note the use of the `right` argument so that the classes are closed on
the left and opened on the right. 


`descstat::recut` can be used to reduce the number of classes, by
providing a numerical vector which should be a subset of the initial
class limits:

```{r }
wages %>% pull(wage) %>% levels %>% head
wages2 <- wages %>% mutate(wage = recut(wage, c(10, 20, 50)))
wages2 %>% pull(wage) %>% levels
```

The `cls2val` function takes as arguments a series which contains
numerical classes, a `pos` argument and the three arguments `xfirst`,
`xlast` and `wlast` previously described. `pos` is a numerical value
which can take any value between 0 and 1:

- `pos = 0` returns the lower bound of the class,
- `pos = 1` returns the upper bound of the class,
- `pos = 0.5` returns the center of the class. 

```{r }
wages2 %>% select(wage) %>%
    mutate(lb = cls2val(wage, 0),
           ub = cls2val(wage, 1),
           ct = cls2val(wage, 0.5, xfirst = 5, xlast = 100))
```

## Ploting bins tables


`ggplot` provides three geoms to plot the distribution of a numerical
series: `geom_histogram`, `geom_density` and `geom_freqpoly`. These
three geoms use the `bin` stat, which means that they consider a raw
vector of numerical values, create classes, count the number of
observations in each class and then plot the result:


```{r }
padova %>% ggplot(aes(price)) +
    geom_histogram(aes(y = ..density..), color = "black", fill = "white") +
    geom_freqpoly(aes(y = ..density..), color = "red") + geom_density(color = "blue")
```

These geoms can be used when individual numerical data are available,
but not when only classes are observed. A `pre_plot` method is
provided for `bins_table` objects, with the `plot` argument either
equal to `histogram` (the default) or `freqpoly`. The resulting table
contains columns called `x` and `y` and can be ploted using
`geom_polygon` for an histogram and `geom_line` for a frequency
polygon.

```{r }
wages %>% bins_table(wage, "d", breaks = c(10, 20, 30, 40, 50)) %>%
    pre_plot(plot = "histogram") %>%
    ggplot(aes(x, y)) + geom_polygon(fill = "white", color = "black")
wages %>% bins_table(wage, "d", breaks = c(10, 20, 30, 40, 50)) %>%
    pre_plot(plot = "freqpoly") %>%
    ggplot(aes(x, y)) + geom_line()
```

Lorenz curve can be plotted using `plot = "lorenz"`. Not that in this
case, the table should contain `F` and `M`.


```{r }
lzc <- wages %>% bins_table(wage, "MF", breaks = c(10, 20, 30, 40, 50)) %>%
    pre_plot(plot = "lorenz")
lzc
```
Each line in the resulting tibble indicate the coordinates (`F` for
`x` and `M` for `y`) of the points that are necessary to plot the
polygons under the Lorenz curve. `pts` is a logical which indicates
which lines correspond to points that are on the Lorenz 



```{r }
lzc %>% ggplot(aes(F, M)) +
    geom_polygon(fill = "lightyellow", color = "black") +
    geom_point(data = filter(lzc, pts)) +
    geom_line(data = tibble(F = c(0, 1), M = c(0, 1)), color = "blue") +
    geom_line(data = tibble(F = c(0, 1, 1), M = c(0, 0, 1)), color = "red")
```


## Computing descriptive statistics

Descriptive statistics can easily be computed applying functions to
`bins_table` objects. The problem is that only a few statistical
functions of the `base` and the `stats` packages are generic. For
these, methods where written for `bins_table` objects, for the other
ones, we had to create new functions.

```{r echo = FALSE}
tribble(~ "base R", ~ descstat,
        "mean", "mean",
        "median", "median",
        "quantile", "quantile",
        "var", "variance",
        "sd", "stdev",
        "mad", "madev",
        "", "modval",
        "", "medial",
        "", "gini") %>%
    knitr::kable()
```

To compute the central values statistics of the distribution of
`wages` we use:

```{r collapse = TRUE}
z <- wages %>% bins_table(wage)
z %>% mean
z %>% median
z %>% modval
```

`median` returns a value computed using the Thales theorem. `modval`
returns the mode, which is a one line tibble containing the class, the
center of the class and the modal value.

For the dispersion statistics:

```{r collapse = TRUE}
z %>% stdev
z %>% variance
```
For the quantiles, the argument `y` can be used to compute the
quantiles using the values of the variable (`y = "value"`, the
default) or the masses (`y = "mass"`):


```{r collapse = TRUE}
z %>% quantile(probs = c(0.25, 0.5, 0.75))
z %>% quantile(y = "mass", probs = c(0.25, 0.5, 0.75))
```

The quantile of level 0.5 is the median if the first case, the medial
in the second case

```{r collapse = TRUE}
z %>% median
z %>% medial
```

`gini` computes the Gini coefficient of the series:


```{r }
z %>% gini
```

# Contingency tables


With `dplyr`, a contingency table can be computed using `count` with
two categorical variables. 

Let's first reduce the number of classes of `size` and `wage` in the
`wages` table.

```{r }
wages2 <- wages %>%
    mutate(size = recut(size, c(20, 50, 100)),
           wage = recut(wage, c(10, 30, 50)))
```

```{r }
wages2 %>% count(size, wage)
```

To get a "wide" table, with the values of one of the two variables
being the columns, we can use `tidyr::pivot_wider`:


```{r }
wages2 %>% count(size, wage) %>%
    tidyr::pivot_wider(values_from = n, names_from = size)
```
## Creating contigency tables using `cont_table`

The same contingency table can be obtained using
`descstat::cont_table`:

```{r }
wages2 %>% cont_table(wage, size)
```

The result is a `cont_table` object, which is a tibble in "long"
format, as the result of the `dplyr::count` function. The printing of
the table in "wide" format is performed by the `pre_print` method,
which is included in the `format` method, but should be used
explicitely while using `knitr::kable`.

```{r }
wages2 %>% cont_table(wage, size) %>%
    pre_print %>% knitr::kable()
```

The `total` argument can be set to `TRUE` to get a row and a column of
totals for the two series.


```{r }
wages2 %>% cont_table(wage, size, total = TRUE)
```

A `weights` argument is used to mimic the population. For example, the
`employment` table contains a series of weights called `weights`:


```{r }
employment %>% cont_table(age, sex, weights = weights, total = TRUE)
```

as for `bins_table`, central values for the first and the last class
can be set using arguments `xfirst#`, `xlast#` and `wlast#`, where `#`
is equal to 1 or 2 for the first and the second variable indicated in
the `cont_table` function.


## Plotting a contingency table

A contingency table can be ploted using `geom_point`, with the size of
the points being proportional to the count of the cells. The `pre_plot`
method replaces classes by values.

```{r }
wages2 %>% cont_table(size, wage) %>% pre_plot %>%
    ggplot() + geom_point(aes(size, wage, size = n))
```


## Computing the distributions from a contingency table

$n_{ij}$ is the count of the cell corresponding to the $i$^th^
modality of the first variable and the $j$^th^ modality of the second
one.

- the joint distribution is obtained by dividing $n_{ij}$ by the
  sample size,
- the marginal distribution is obtained by summing the counts
  column-wise $n_{i.}=\sum_{j}n_{ij}$ for the first variable and
  row-wise $n_{.j} = \sum_{i}n_{ij}$ for the second one,
- the conditional distribution is obtained by dividing the joint by
  the marginal frequencies.
  
The `joint`, `marginal` and `conditional` functions return these three
distributions. The last two require an argument `y` which is one of the
two variables of the `bins_table` object.

```{r }
wht <- wages2 %>% cont_table(size, wage)
wht %>% joint
wht %>% marginal(size)
wht %>% conditional(size)
```

## Computing descriptive statistics

Descriptive statistics can be computed using any of the three
distributions. Using the joint distribution, we get a tibble
containing two columns for the two variables

```{r }
wht %>% joint %>% mean
wht %>% joint %>% stdev
wht %>% joint %>% variance
```
The same (univariate) statistics can be obtained using the marginal
distribution:

```{r }
wht %>% marginal(size) %>% mean
wht %>% marginal(size) %>% stdev
```
or even more simply considering the univariate distribution computed
by `bins_table`:

```{r }
wages2 %>% bins_table(size) %>% mean
```

The `mean`, `stdev` and `variance` methods are actually only usefull
when applied to a conditional distribution; in this case, considering
for example the conditional distribution of the first variable, there
are as many values returned that the number of modalities of the
second (conditioning) variable.

```{r }
wht %>% conditional(wage) %>% mean
wht %>% conditional(wage) %>% variance
```

## Regression curve


The total variance of $X$ can be writen as the sum of:

- the explained variance, *ie* the variance of the conditional means,
- the residual variance, *ie* the mean of the conditional variances.

$$
s_{x}^2 = \sum_j f_{.j} s^2_{x_j} + \sum_j f_{.j} (\bar{x}_j -
\bar{\bar{x}}) ^ 2
$$

The decomposition of the variance can be computed by joining tables
containing the conditional moments and the marginal distribution of
the conditioning variable and then applying the formula:

```{r }
cm <- wht %>% conditional(wage) %>% mean %>% rename(mean = wage)
cv <- wht %>% conditional(wage) %>% variance %>% rename(variance = wage)
md <- wht %>% marginal(size)
md %>% left_join(cm) %>% left_join(cv) %>%
    summarise(om = mean(mean),
              ev = sum(f * (mean - om) ^ 2),
              rv = sum(f * variance),
              tv = ev + rv)                       
```

Or more simply using the `var_decomp` function:

```{r }
wht_wage <- wht %>% var_decomp("wage")
wht_wage
```
which has a `summary` method which computes the different elements of
the decomposition:

```{r }
wht_wage %>% summary
```
and especially the correlation ratio, which is obtained by dividing
the explained variance by the total variance. 

The regression curve of `wage` on `size` can be plotted using
`wht_wage`, together with error bars:

```{r }
wht_wage %>% ggplot(aes(size_val, mean)) + geom_point() +
    geom_line(lty = "dotted") +
    geom_errorbar(aes(ymin = mean - sqrt(var), ymax = mean + sqrt(var))) +
    labs(x = "size", y = "wage")    
```

## Linear regression


For the joint distribution, two other functions are provided,
`covariance` and `correlation` (which are the equivalent of the
non-generic `stats::cov` and `stats::cor` functions) to compute the
covariance and the linear coefficient of correlation.

```{r }
wht %>% joint %>% covariance
wht %>% joint %>% correlation
```

The regression line can be computed using the `regline` function:


```{r }
rl <- regline(wage ~ size, wht)
rl
```
which returns the intercept and the slope of the regression of `wage`
on `size`. We can then draw the points and the regression line:


```{r }
wht %>% pre_plot %>% ggplot() + geom_point(aes(size, wage, size = n)) +
    geom_abline(intercept = rl[1], slope = rl[2])
```
